Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-04-04 manuscript/main.tex
small titile page edits

\author[1]{Andrew J.~Holbrook}
\author[1]{Andrew F.~Magee}
\author[2]{Nathan E.~Glatt-Holtz}
\author[1,3,4]{Marc A.~Suchard}

 \affil[1]{Department of Biostatistics, University of California, Los Angeles}
 \affil[2]{Department of Mathematics, Tulane University}
 \affil[3]{Department of Biomathematics, University of California, Los Angeles}
 \affil[4]{Department of Human Genetics, University of California, Los Angeles}

@@

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.  Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.  But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.   \textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.  Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1,900 model parameters.

@@

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\bigg)^{\!T}\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \Biggl(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \Biggr)^{\!T} (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.

>>>>

\author[1]{Gustavo Didier}
\author[1]{Nathan E.~Glatt-Holtz}
\author[2]{Andrew J.~Holbrook\thanks{Corresponding author: \url{aholbroo@g.ucla.edu}.}}
\author[2]{\\Andrew F.~Magee}
\author[2,3,4]{Marc A.~Suchard}

 \affil[1]{Department of Mathematics, Tulane University}
  \affil[2]{Department of Biostatistics, University of California, Los Angeles}
 \affil[3]{Department of Biomathematics, University of California, Los Angeles}
 \affil[4]{Department of Human Genetics, University of California, Los Angeles}

@@

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.  Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.  But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.   \citet{magee2023random} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.  Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1,900 model parameters.

@@

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta) 
	&=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial (\p_v^T\q_v)}{\partial \P_v} \frac{\partial \P^T_v}{\partial \theta_d} \right)
	= \sum_{v=1}^{2N-2} \mbox{tr} \left(  (\q_u\circ \P_w\p_w) \p_v^T\bigg( \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d}\bigg)^{\!T}\right)  \\ \nonumber
	&=\sum_{v=1}^{2N-2} \p_v^T \Biggl(\sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \Biggr)^{\!T} (\q_u\circ \P_w\p_w)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \citep{magee2023random} and specifies
\begin{align}\label{eq:mixed}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \citet{magee2023random} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.
